pSEOBNRv4HM

### 2020-06-30 papers/intro_paper/intro_paper.tex
updated the draft with comments and references

\documentclass[twocolumn,prd,superscriptaddress,amsfonts,amssymb,amsmath,preprintnumbers]{revtex4-1}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{bm}
\usepackage[dvipsnames]{xcolor}
\usepackage{bm}
\usepackage{times}
\usepackage{xspace}
\usepackage[varg]{txfonts}
\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}

@@

The LIGO Scientific Collaboration~\citep{LSC} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave observations~\citep{GWTC-2} from the first half of their third observing run~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{GWTC-1}, the Advanced LIGO detectors at Hanford, Washington~\citep{technicaldocument_AdL_H} and Livingston, Louisiana~\citep{technicaldocument_AdL_L}, and the Advanced Virgo detector in Cascina, Italy~\citep{technicaldocument_V} have now detected more than $40$ gravitational wave events from the merger of compact objects like neutron stars and/or black holes (compact binary coalescences/CBCs). Alongside independent claims of detections~\citep{Nitz,IAS}, this has firmly established the field of gravitational wave astronomy, five years on from the first ever direct detection of gravitational waves on Earth, on September 14, 2015~\citep{GW150914det}. 

@@

Observation of gravitational waves have had significant astrophysical and cosmological implications~\citep{LVC:astropapers,H0paper}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{GW150914TGR,GW170817TGR,GWTC-1TGR}. In GR, a CBC involving two black holes (a binary black hole/BBH system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} event marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  We have performed tests of gravitational wave generation and source dynamics, where we place bounds on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of gravitational wave propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{GW170104det,samajdaretal} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{GW170814det,isietal}. We have also checked for consistency between different portions of the signal using estimates of final mass and spin~\citep{IMRCTpapers}, and consistency of the residuals with interferometric noise~\citep{sudarshanetal,GWTC-1TGR}. None of these tests report any departure from the predictions of GR.

@@

Tests of black hole ringdown and nature of the remnant object is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as a theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{GW150914TGR}, and later extended to include overtones in~\citep{isietal}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{carulloetalpapers,bertipapers}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Miriametal} or through search for echos in the post-merger signal~\citep{nielsen,abedi,kawa}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Most of the tests mentioned above focus on analysing the post-merger or late-time ringdown signal in isolation. Second generation ground-based interferometric detectors like Advanced LIGO and Virgo are most sensitive to stellar-mass black hole binaries that merge near the minima of their sensitivity band ($\sim 100$Hz). As a consequence, the remnant object \textit{rings down} in shot-noise dominated higher frequencies, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. Furthermore, the post-merger signal of a BBH coalescence transitions from a non-linear to a linear regime where the description of the signal as a QNM spectrum becomes appropriate~\citep{}. The ringdown start time isn't a clearly defined quantity and has been explored in detail in~\citep{bhagwatetal}. In other works~\citep{carulloetal}, it has been left as a free parameter to be estimated directly from the data. Uncertainties in estimates of the ringdown start-time, as well as an overall lack of SNR in the post-merger signal, given typical sensitivities of ground-based detectors result in significant statistical uncertainties in the measurement of the QNM frequencies.

@@

An independent approach to black hole spectroscopy, based on a full-signal analysis, was outlined in~\citep{britopaper1} (henceforth referred to as \paperone). \abhi{describe underlying GR model; parameterisation; advantages over other methods; highlight spin}

@@

\section{Waveform Model}\label{sec:model}
SEOBNRv4HM model + parameterised
highlight importance of spins

\section{Method/Bayesian PE}\label{sec:method}
Bayesian inference

\section{Results}\label{sec:results}
Priors
LIGO-Virgo results
Results on injections: modGR, noise systematics

@@

>>>>

\documentclass[twocolumn,prd,superscriptaddress,amsfonts,amssymb,amsmath,preprintnumbers]{revtex4-1}
\usepackage{epsfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{bm}
\usepackage[dvipsnames]{xcolor}
\usepackage{bm}
\usepackage{times}
\usepackage{xspace}
\usepackage[varg]{txfonts}
\usepackage[normalem]{ulem} % To get strikethrough (\sout)
\usepackage[colorlinks]{hyperref}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{url}

@@

The LIGO Scientific Collaboration~\citep{lsc} and the Virgo Collaborations~\citep{Virgo} recently announced their catalogue of gravitational wave observations~\citep{GWTC-2} from the first half of their third observing run~\citep{O3reference}. Combined with the confirmed observations from the first and second observing runs~\citep{abbott2019gwtc}, the Advanced LIGO detectors at Hanford, Washington and Livingston, Louisiana~\citep{aasi2015characterization}, and the Advanced Virgo detector in Cascina, Italy~\citep{acernese2014advanced} have now detected more than $40$ gravitational wave events from the merger of compact objects like neutron stars and/or black holes (compact binary coalescences/CBCs). Alongside independent claims of detections~\citep{nitz20191,nitz20202,2019PhRvD.100b3007Z,2020PhRvD.101h3030V,Venumadhav_2020}, this has firmly established the field of gravitational wave astronomy, five years on from the first ever direct detection of gravitational waves on Earth, on September 14, 2015~\citep{abbott2016observation}. 

@@

Observation of gravitational waves have had significant astrophysical and cosmological implications~\citep{LSC_2016astroph,gw170817_mma,gw170817_joint,gw170817_hubble}. It has also allowed us to make statements in fundamental physics. Specifically, LIGO-Virgo's observations have allowed us to test predictions of Einstein's theory of General Relativity~\citep[GR]{}, in previously unexplored regimes of highly relativistic, strong-field regimes of gravity~\citep{LSC_2016grtests,GW170817_TGR,gwtc1_tgr}. In GR, a CBC involving two black holes (a binary black hole/BBH system) is described in three distinct phases: an early \textit{inspiral}, where the two compact objects spiral in and \textit{plunge} due to a back-reaction of gravitational-wave-emission, a \textit{merger} event marked by the formation of an apparent horizon~\citep{NRpaper}, and a late-time \textit{ringdown}, where the newly formed remnant object settles down to a stable Kerr state through the emission of an exponentially damped quasi-normal-mode (QNM) spectrum of gravitational radiation~\citep{vishu,earlyqnmpapers}.  We have performed tests of gravitational wave generation and source dynamics, where we place bounds on parametrised deviations in the Post-Newtonian coefficients describing the early inspiral~\citep{earlydevelopmentpapers}, and phenomenological coefficients describing the intermediate (plunge) and merger regimes of coalescence~\citep{TIGERmethodspapers}, tests of gravitational wave propagation, which assume a generalised dispersion relation and place upper bounds on the Compton wavelength and consequently, the mass of the graviton~\citep{gw170104,samajdar2017projected} and tests of the polarisation of gravitational radiation using a multi-gravitational-wave-detector network~\citep{gw170814,isi2017probing}. We have also checked for consistency between different portions of the signal using estimates of final mass and spin~\citep{Ghosh:2016xx,Ghosh:2017gfp,LSC_2016grtests}, and consistency of the residuals with interferometric noise~\citep{Ghonge:2020suv,gwtc1_tgr}. None of these tests report any departure from the predictions of GR.

@@

Tests of black hole ringdown and nature of the remnant object is an active field of research at the moment. The no-hair conjecture in GR~\citep{} states that an (electrically neutral) astrophysical black hole is completely described by two observables: mass and spin angular momentum. One consequence of the no-hair conjecture is that the (complex) QNM frequencies of gravitational radiation emitted by a perturbed isolated black hole is uniquely determined by its mass and spin angular momentum. Hence a test of the no-hair conjecture would involve checking for consistency between estimates of mass and spin of the remnant object across multiple QNM frequencies. An inconsistency would either indicate a non-black hole nature of the remnant object, or an incompleteness of GR as a theory of gravity. The consistency between the post-merger signal and the least dampled QNM was first demonstrated in ~\citep{LSC_2016grtests}, and later extended to include overtones in~\citep{Giesler:2019uxc,Isi:2019aib,Bhagwat:2019dtm,Forteza:2020hbw}. Consistency of the late-time signal with a single QNM is a test of the ringdown of a BBH coalescence, but not necessarily a test of the no-hair conjecture, which requires the measurement of (atleast) two QNMs (black hole spectroscopy), and checking for consistency between them. Recent work in that direction include~\citep{Carullo:2018gah,Carullo:2019flw,Bhagwat:2019bwv}. The nature of the remnant object has also been explored through tests of black hole thermodynamics, like the Hawking's area theorem~\citep{Cabero:2017avf} or through search for echos in the post-merger signal~\citep{Nielsen:2018lkf,Tsang:2019zra,Lo:2018sep,Abedi:2018npz,Abedi:2020sgg,Testa:2018bzd}. None of these tests have found evidence for non-black hole nature of the remnant object (as described in GR) in LIGO-Virgo BBH observations.

@@

Most of the tests mentioned above focus on analysing the post-merger or late-time ringdown signal in isolation. Second generation ground-based interferometric detectors like Advanced LIGO and Virgo are most sensitive to stellar-mass black hole binaries that merge near the minima of their sensitivity band ($\sim 100$Hz). As a consequence, the remnant object \textit{rings down} in shot-noise dominated higher frequencies, leaving very little signal-to-noise ratio (SNR) in the post-merger signal. Furthermore, the post-merger signal of a BBH coalescence transitions from a non-linear to a linear regime where the description of the signal as a QNM spectrum becomes appropriate~\citep{}. The ringdown start time isn't a clearly defined quantity and has been explored in detail in~\citep{Bhagwat:2017tkm}. In other works~\citep{Carullo:2018gah,Carullo:2019flw}, it has been left as a free parameter to be estimated directly from the data. Uncertainties in estimates of the ringdown start-time, as well as an overall lack of SNR in the post-merger signal, given typical sensitivities of ground-based detectors result in significant statistical uncertainties in the measurement of the QNM frequencies.

@@

An independent approach to black hole spectroscopy, based on a full-signal analysis, was outlined in~\citep{Brito:2018rfr} (henceforth referred to as \paperone). \abhi{describe underlying GR model; parameterisation; advantages over other methods; highlight spin}

@@

\section{Waveform Model}\label{sec:model}

\abhi{Summary of the section:}

\begin{itemize}
\item{a description of the SEOBNRv4HM model and it's generalisation to form the pSEOBNRv4HM model}
\item{highlight the importance of spins in the model, especially with regards to the non-spinning method outlined in Brito et al.}
\end{itemize}

\section{Method/Bayesian PE}\label{sec:method}

\abhi{This section will basically provide an overview of Bayesian inference. Starting from the generalised definition of likelihood, provide an expression for the posterior on the delta QNM parameters.}

\section{Results}\label{sec:results}
\subsection{Priors}

\abhi{A small sub-section dedicated to priors used in this analysis. Very common question from referees.}

\subsection{Results on GW150914-like injection}

\abhi{The GW150914-like injections we have:}
\begin{itemize}
\item{an SEOBNRv4HM software injection with the 22 mode frequencies estimated}
\item{an SEOBNRv4HM software injection with the 33 mode frequencies estimated}
\item{a non-spinning NR injection with parameters close to the real event but higher SNR for which 22 mode frequencies are estimated}
\item{a non-spinning highly-asymmetric (q=6) NR injection with total mass similar to GW150914 and high SNR for which 22 and 33 mode frequencies are estimated}
\end{itemize}

\subsection{Results on modGR injection}

\abhi{As part of review we had developed the infrastructure to make non-GR waveforms using pSEOBNRv4HM, and running LALInference on it. We had specifically chosen an injection with domega220 = dtau220  = 0.5 in zero-noise, and successfully recovered those frequencies thus demonstrating the robustness of the method in detecting deviations from GR.}

\subsection{Results on actual LIGO-Virgo results}

\abhi{Results obtained as part of the review on GW150914 and other O1-O2 events which are of relevance}

\subsection{Results on noise systematics study}

\abhi{A small summary of the noise systematics study that was done for S190521g to show that the deviations seen were consistent with systematics due to noise. The study include 2 separate sections: in Gaussian noise, and in real LIGO-Virgo noise immediately adjacent to the actual event.}

@@

%
\bibliographystyle{apsrev}
\bibliography{intro_paper}
